ppseq: An R Package for Sequential Predictive Probability Monitoring

### 2022-01-25 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
fixing issues with html paper

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent successes for non-cytotoxic treatments have relied on expansive early phase clinical trials that far exceeded the size of trials typically implemented for chemotherapies, with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements. With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential futility monitoring based on the Bayesian predictive probability. Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib

---

@@

The \CRANpkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility. The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial. The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error. The \code{calibrate\_thresholds()} function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument \code{pp\_threshold} for posterior thresholds and the argument \code{ppp\_threshold} for predictive thresholds. Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}. Default arguments are available for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}, but customization is also possible. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design. The function returns a list, the first element of which is a \code{tibble} containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative. The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power. The \code{print()} option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.

@@

After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options. The \CRANpkg{ppseq} package introduces two optimization criteria to assist users in making a selection. The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the type I error by the power. The second, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the average sample size under the null by the average sample size under the alternative, subject to constraints on the type I error and power. The \code{optimize\_design()} function will return a list that contains the details of each of the two optimal designs.

@@

To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function. The function takes the sample sizes \code{n} at which interim analyses are to be performed as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0} (set to \code{NULL} in the two-sample case), the posterior threshold of the selected design \code{theta}, and the predictive threshold of the selected design \code{ppp}. Default arguments are provided for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}, though customization is also possible. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to \code{NULL} in the one-sample case). The function results in a \code{tibble}. In the one-sample case, the \code{tibble} includes a column for the sample size \code{n} at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision \code{ppp}. The trial would stop at a given look if the number of observed responses is less than or equal to $r$ otherwise the trial would continue enrolling if the number of observed responses is greater than $r$. At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is greater than $r$. In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively. There are also columns for \code{r0} and \code{r1}, the number of responses in the control arm and experimental arm leading to the decision, respectively. Finally, there is a column for the posterior predictive probability associated with that decision \code{ppp}.

@@

Finally, to assist users in comparing the results of the various design options, a \code{plot()} option is available for the results of \code{calibrate\_thresholds} that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020]. Two plots are produced, one plotting type I error by power and indicating the optimal accuracy design, and one plotting the average sample size under the null by the average sample size under the alternative and indicating the optimal efficiency design. The motivation for including an interactive graphics option was the utility of the additional information available when hovering over each point. Instead of simply eyeballing where points fall along the axes, users can see the specific type I error, power, average sample size under the null, average sample size under the alternative, the posterior and predictive thresholds associated with the design, as well as the distance to the upper left point on the plot. A \code{plot()} option is also available for the results of \code{calc\_decision\_rules}. In the one-sample case it produces a single plot showing the sample size at each interim analysis on the x-axis and the possible number of responses at each interim analysis on the y-axis. In the two-sample case a grid of plots is produced, with one plot for each interim analysis. The x-axis shows the number of possible responses in the control group and the y-axis shows the number of possible responses in the experimental group. In both cases, the boxes are colored green for a "proceed" decision and red for a "stop" decision for each combination and the hover box produced by \CRANpkg{plotly} provides the details.

@@

We use the \code{calibrate\_thresholds()} function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds. Because of the inherent computation intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations. The user will be responsible for setting up a call to \code{future::plan()} that is appropriate to their operating environment and simulation setting. The example in this case study was run on a Unix server with 192 cores, and we wished to use 76 cores to accommodate the 76 distinct designs that result from the 19 posterior by 4 predictive threshold grid. Because the code takes some time to run, the results of the below example code are available as a dataset called \code{one\_sample\_cal\_tbl} included in the \CRANpkg{ppseq} package.

@@

Next we print the results table using the \code{print()} option, and limited to designs with type I error between 0.05 and 0.1, and a minimum power of 0.7. We find that 35 of the 76 designs meet these criteria for type I error and power.

@@

We use the \code{optimize\_design()} function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power. We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05. It has a type I error of 0.072, power of 0.883, average sample size under the null of 51, and average sample size under the alternative of 91. The optimal efficiency design is the one with posterior threshold of 0.92 and predictive threshold of 0.1. It has a type I error of 0.06, power of 0.796, average sample size under the null of 39, and average sample size under the alternative of 82. For comparison, the original design of the atezolizumab expansion cohort in mUC, with a single look for futility after the first 14 patients, has a type I error of 0.005, power of 0.528, average sample size under the null of 76, and average sample size under the alternative of 92.

@@

To compare these optimal designs with all other designs, we can use the \code{plot()} function with the \code{plotly = TRUE} option to obtain interactive visualizations, or the \code{plotly = FALSE} option to obtain static visualizations.

@@

In this case we may choose to use the optimal efficiency design, which has the desirable trait of a very small average sample size under the null of just 39 patients, while still maintaining reasonable type I error of 0.06 and power of 0.796. This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment. Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial. Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \CRANpkg{ppseq} package. In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial. After the first 10 patients we would stop the trial if there were 0 responses, and so on. At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were greater than or equal to 14 responses.

>>>>

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent successes for non-cytotoxic treatments have relied on expansive early phase clinical trials that far exceeded the size of trials typically implemented for chemotherapies, with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements. With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential futility monitoring based on the Bayesian predictive probability. Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib
---

@@

The \CRANpkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility. The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial. The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error. The `calibrate_thresholds()` function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument `pp_threshold` for posterior thresholds and the argument `ppp_threshold` for predictive thresholds. Other required arguments include the null response rate `p_null`, the alternative response rate `p_alt`, a vector of sample sizes at which interim analyses are to be performed `n` as well as the maximum total sample size `N`. Default arguments are available for the direction of the alternative hypothesis `direction`, a vector of the two hyperparameters of the prior beta distribution `prior`, and the number of posterior samples `S` and the number of simulated trial datasets `nsim`, but customization is also possible. The additional argument `delta` can specify the clinically meaningful difference between groups in the case of a two-sample trial design. The function returns a list, the first element of which is a `tibble` containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative. The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power. The `print()` option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.

@@

After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options. The \CRANpkg{ppseq} package introduces two optimization criteria to assist users in making a selection. The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the type I error by the power. The second, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the average sample size under the null by the average sample size under the alternative, subject to constraints on the type I error and power. The `optimize_design()` function will return a list that contains the details of each of the two optimal designs.

@@

To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the `calc_decision_rules()` function. The function takes the sample sizes `n` at which interim analyses are to be performed as well as the maximum total sample size `N`, the null value to compare to in the one-sample case `p0` (set to `NULL` in the two-sample case), the posterior threshold of the selected design `theta`, and the predictive threshold of the selected design `ppp`. Default arguments are provided for the direction of the alternative hypothesis `direction`, a vector of the two hyperparameters of the prior beta distribution `prior`, and the number of posterior samples `S`, though customization is also possible. The additional argument `delta` can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to `NULL` in the one-sample case). The function results in a `tibble`. In the one-sample case, the `tibble` includes a column for the sample size `n` at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision `ppp`. The trial would stop at a given look if the number of observed responses is less than or equal to $r$ otherwise the trial would continue enrolling if the number of observed responses is greater than $r$. At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is greater than $r$. In the two-sample case, the `tibble` includes columns for `n0` and `n1`, the sample size at each interim analysis in the control and experimental arms, respectively. There are also columns for `r0` and `r1`, the number of responses in the control arm and experimental arm leading to the decision, respectively. Finally, there is a column for the posterior predictive probability associated with that decision `ppp`.

@@

Finally, to assist users in comparing the results of the various design options, a `plot()` option is available for the results of `calibrate_thresholds` that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020]. Two plots are produced, one plotting type I error by power and indicating the optimal accuracy design, and one plotting the average sample size under the null by the average sample size under the alternative and indicating the optimal efficiency design. The motivation for including an interactive graphics option was the utility of the additional information available when hovering over each point. Instead of simply eyeballing where points fall along the axes, users can see the specific type I error, power, average sample size under the null, average sample size under the alternative, the posterior and predictive thresholds associated with the design, as well as the distance to the upper left point on the plot. A `plot()` option is also available for the results of `calc_decision_rules`. In the one-sample case it produces a single plot showing the sample size at each interim analysis on the x-axis and the possible number of responses at each interim analysis on the y-axis. In the two-sample case a grid of plots is produced, with one plot for each interim analysis. The x-axis shows the number of possible responses in the control group and the y-axis shows the number of possible responses in the experimental group. In both cases, the boxes are colored green for a "proceed" decision and red for a "stop" decision for each combination and the hover box produced by \CRANpkg{plotly} provides the details.

@@

We use the `calibrate_thresholds()` function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds. Because of the inherent computation intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations. The user will be responsible for setting up a call to `future::plan()` that is appropriate to their operating environment and simulation setting. The example in this case study was run on a Unix server with 192 cores, and we wished to use 76 cores to accommodate the 76 distinct designs that result from the 19 posterior by 4 predictive threshold grid. Because the code takes some time to run, the results of the below example code are available as a dataset called `one_sample_cal_tbl` included in the \CRANpkg{ppseq} package.

@@

Next we print the results table using the `print()` option, and limited to designs with type I error between 0.05 and 0.1, and a minimum power of 0.7. We find that 35 of the 76 designs meet these criteria for type I error and power.

@@

We use the `optimize_design()` function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power. We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05. It has a type I error of 0.072, power of 0.883, average sample size under the null of 51, and average sample size under the alternative of 91. The optimal efficiency design is the one with posterior threshold of 0.92 and predictive threshold of 0.1. It has a type I error of 0.06, power of 0.796, average sample size under the null of 39, and average sample size under the alternative of 82. For comparison, the original design of the atezolizumab expansion cohort in mUC, with a single look for futility after the first 14 patients, has a type I error of 0.005, power of 0.528, average sample size under the null of 76, and average sample size under the alternative of 92.

@@

To compare these optimal designs with all other designs, we can use the `plot()` function with the `plotly = TRUE` option to obtain interactive visualizations, or the `plotly = FALSE` option to obtain static visualizations.

@@

In this case we may choose to use the optimal efficiency design, which has the desirable trait of a very small average sample size under the null of just 39 patients, while still maintaining reasonable type I error of 0.06 and power of 0.796. This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment. Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial. Because of the computational time involved, the results of the below example code are available as a dataset called `one_sample_decision_tbl` included in the \CRANpkg{ppseq} package. In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial. After the first 10 patients we would stop the trial if there were 0 responses, and so on. At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were greater than or equal to 14 responses.
